---
title: "Homework 6, Luke Flyer"
format: 
  pdf:
    fig-height: 4
    fig-width: 6
output: pdf_document
---

```{r, echo = FALSE, warning = FALSE}
#| label: load-packages

library(tidyverse)
library(broom)
library(knitr)

# add other packages as needed
```

## Data: Association Between Bird-Keeping and Risk of Lung Cancer {.midi}

A 1972-1981 health survey in The Hague, Netherlands, discovered an association between keeping pet birds and increased risk of lung cancer. To investigate birdkeeping as a risk factor, researchers conducted a case-control study of patients in 1985 at four hospitals in The Hague. They identified 49 cases of lung cancer among patients who were registered with a general practice, who were age 65 or younger, and who had resided in the city since 1965. Each patient (case) with cancer was matched with two control subjects (without cancer) by age and sex. Further details can be found in Holst, Kromhout, and Brand (1988).

Age, sex, and smoking history are all known to be associated with lung cancer incidence. Thus, researchers wished to determine after age, sex, socioeconomic status, and smoking have been controlled for, is an additional risk associated with birdkeeping? The data (Ramsey and Schafer 2002) is found in $$`birdkeeping.csv`(data/birdkeeping.csv)$$.\footnote{This problem is adapted from Section 6.8.1, Ex 4.}

The paper that this exercise is based upon can be found here and please read it before completing the assignment. (<https://www.bmj.com/content/bmj/297/6659/1319.full.pdf>)

# Exercise 1

::: callout-note
## Part a

Create a segmented bar chart and appropriate table of proportions showing the relationship between birdkeeping and cancer diagnosis. Summarize the relationship in 1 - 2 sentences.
:::

```{r}

df <- read.csv("birdkeeping.csv")

table <- table(df$bird, df$cancer)

table_prop <- prop.table(table, 1)

table_prop_df <- as.data.frame(table_prop)

colnames(table_prop_df) <- c("Birdkeeping", "Cancer Diagnosis", "Proportion")

table_prop_wide <- reshape(table_prop_df, 
                           idvar = "Birdkeeping", 
                           timevar = "Cancer Diagnosis", 
                           direction = "wide")

colnames(table_prop_wide) <- c("Birdkeeping", "No Cancer (0)", "Cancer (1)")

kable(table_prop_wide, format = "markdown", caption = "Proportions of Cancer Diagnosis by Birdkeeping Status")
```

```{r}

ggplot(as.data.frame(table_prop), aes(x = factor(Var1), y = Freq, fill = factor(Var2))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Birdkeeping Status", y = "Proportion", fill = "Cancer Diagnosis") +
  scale_x_discrete(labels = c("No Birdkeeping", "Birdkeeping")) +
  scale_fill_manual(values = c("lightblue", "salmon"), labels = c("No Cancer", "Cancer")) +
  ggtitle("Proportion of Cancer Diagnosis by Birdkeeping Status") +
  theme_minimal()
```

From the above table and segmented bar chart, we can see that there is a higher proportion of lung cancer diagnosis for the birdkeepers than for the non-birdkeepers. While only 20% of non-birdkeepers were diagnosed with cancer, almost 50% of birdkeepers were diagnosed with cancer.

::: callout-note
## Part b

Calculate the unadjusted odds ratio of a lung cancer diagnosis comparing birdkeepers to non-birdkeepers. Interpret this odds ratio in context. (Note: an unadjusted odds ratio is found by not controlling for any other variables.)
:::

```{r}

unadjust <- glm(cancer ~ bird, data = df, family = binomial)
tidy(unadjust) |>
kable(digits = 3)
```

The odds of a lung cancer diagnosis for non-birdkeepers is $e^{-1.386} = 0.250$. The odds of a lung cancer diagnosis for birdkeepers is $e^{-1.386 +1.356} = 0.970$. Therefore, the unadjusted odds ratio of a lung cancer diagnosis comparing birdkeepers to non-birdkeepers is $0.970 / 0.250 = 3.88$.

::: callout-note
## Part c

Does there appear to be an interaction between number of years smoked and whether the subject keeps a bird? Demonstrate with an appropriate plot and briefly explain your response.
:::

```{r}

ggplot(df, aes(x = factor(bird), y = yrsmoke, color = factor(cancer))) +
  geom_boxplot(alpha = 0.6, outlier.shape = 16) +  # Boxplot with colored boxes
  labs(x = "Birdkeeping Status", y = "Years Smoked", color = "Cancer Diagnosis (0 = No, 1 = Yes)") +
  scale_color_manual(values = c("blue", "red"), labels = c("No Cancer", "Cancer")) +
  theme_minimal() +
  ggtitle("Distribution of Years Smoked by Birdkeeping Status and Cancer Diagnosis") +
  theme(legend.title = element_blank())
```

Based on the boxplots above, it appears as if the affect of years smoked on cancer diagnosis seems about the same for both birdkeepers and non-birdkeepers. In both cases, those who have a higher number of years smoked seem more likely to get cancer. Because of this similarity in effect of years smoked on cancer between the birdkeepers and non-birdkeepers, we can say that there appears to be no interaction effect.

Before answering the next questions, fit logistic regression models in R with cancer as the response and the following sets of explanatory variables:

-   model1 = age, yrsmoke, cigsday, female, highstatus, bird
-   model2 = yrsmoke, cigsday, highstatus, bird
-   model3 = yrsmoke, bird
-   model4 = yrsmoke, bird, yrsmoke:bird

```{r}

model1 <- glm(cancer ~ age + yrsmoke + cigsday + female + highstatus + bird, 
              data = df, family = binomial)

model2 <- glm(cancer ~ yrsmoke + cigsday + highstatus + bird, 
              data = df, family = binomial)

model3 <- glm(cancer ~ yrsmoke + bird, data = df, family = binomial)

model4 <- glm(cancer ~ yrsmoke * bird, data = df, family = binomial)
```

::: callout-note
## Part d

Is there evidence that we can remove age and female from our model? Perform an appropriate test comparing model1 to model2; give a test statistic and p-value, and state a conclusion in context.
:::

```{r}

drop_in_dev <- anova(model1, model2, test = "Chisq")
drop_in_dev
```

We do not have significant evidence (Drop-in-deviance test: $\chi^2 = -2.5257$, $p = 0.2828$) that supports the model that removes the age and female variables.

::: callout-note
## Part e

Carefully interpret each of the four model coefficients (including the intercept) in model4 in context.
:::

```{r}

tidy(model4) |>
  kable(digits = 3)
```

ASK

For a non-birdkeeper who has smoked for 0 years, the log-odds of a lung cancer diagnosis is expected to be -2.999.

For a non-birdkeeper, for every one year increase in years smoked, the log-odds of a lung cancer diagnosis is expected to increase by 0.053.

For a birdkeeper who has smoked 0 years, the log-odds of a lung cancer diagnosis is expected to be -2.999 + 1.179 = -1.82.

For a birdkeeper, for every one year increase in years smoked, the log-odds of a lung cancer diagnosis is expected to increase by 0.053 + 0.009 = 0.062.

::: callout-note
## Part f

If you replaced yrsmoke everywhere it appears in model4 with a mean-centered version of yrsmoke, tell what would change among these elements: the 4 coefficients, the 4 p-values for coefficients, and the residual deviance.
:::

Replacing yrsmoke with its mean-centered version means subtracting the mean of yrsmoke from each observation. This would changes the intercept and its p-value, since now the intercept reflects a patient smoking the average number of years instead of 0. The coefficient and p-value for yrsmoke stay the same because we're just shifting it, not changing its relationship with the response. Since we're using an interaction variable, shifting yrsmoke will affect the p-value and coefficient for the birdkeeping effect on lung cancer. The interaction term and its p-value won't be impacted.

::: callout-note
## Part g

Observe that model3 is a potential final model based on this set of predictor variables. How does the adjusted odds ratio for birdkeeping from model3 compare with the unadjusted odds ratio you found in (b)? Is birdkeeping associated with a significant increase in the odds of developing lung cancer, even after adjusting for other factors?
:::

```{r}

tidy(model3) |>
  kable(digits = 3)
```

For this adjusted model, the odds of a lung cancer diagnosis for non-birdkeepers is $e^{-3.180} = 0.0416$. The odds of a lung cancer diagnosis for birdkeepers is $e^{-3.180 + 1.476} = 0.182$. Therefore, the adjusted odds ratio of a lung cancer diagnosis comparing birdkeepers to non-birdkeepers is $0.182 / 0.0416 = 4.375$. Given that the unadjusted odds ratio was 3.88, these odds ratios are relatively similar.

In this adjusted model 3, birdkeeping is associated with a significant increase in the odds of developing lung cancer, even after adjusting for years smoked.

::: callout-note
## Part h

Discuss the scope of inference in this study. Can we generalize our findings beyond the subjects in this study? Can we conclude that birdkeeping causes increased odds of developing lung cancer? Do you have other concerns with this study design or the analysis you carried out?
:::

Because this data is correlational, we cannot conclude any causality. Also, I don’t think we can apply these findings to the wider population. There are likely other factors, such as other types of pollution or genetics, that affect lung cancer, which aren't included in the study. The sample size is also too small to make broad conclusions. It seems like the participants also weren’t randomly chosen, the sample doesn’t accurately reflect the general population.

# Exercise 2

(Ataman and Sarıyer 2021) use ordinal logistic regression to predict patient wait and treatment times in an emergency department (ED). The goal is to identify relevant factors that can be used to inform recommendations for reducing wait and treatment times, thus improving the quality of care in the ED.

The data include daily records for ED arrivals in August 2018 at a public hospital in Izmir, Turkey. The response variable is Wait time, a categorical variable with three levels:

-   Patients who wait less than 10 minutes
-   Patients whose waiting time is in the range of 10-60 minutes
-   Patients who wait more than 60 minutes

::: callout-note
## Part a

Compare and contrast the proportional odds model with the multinomial logistic regression model. Write your response using 3 - 5 sentences. You can find a brief review of the proportional odds model here: <https://library.virginia.edu/data/articles/fitting-and-interpreting-a-proportional-odds-model> and <https://online.stat.psu.edu/stat504/lesson/8/8.4>
:::

The proportional odds model and multinomial logistic models both use log odds when making interpretations. They also both use a logit link. Both models can have the same number of coefficients, but the proportional odds model assumes a consistent relationship across all categories, whereas the multinomial model allows for more flexibility by treating categories independently. Essentially, it would depend on your research question whether or not you would want the probability of being in a higher category or simply probability of being in one category versus a baseline.

::: callout-note
## Part b

Table 5 in the paper contains the output for the wait time and treatment time models. Consider only the model for wait time. Describe the effect of arrival mode (ambulance, walk-in) on the waiting time. Note: walk-in is the baseline in the model. (A link to the paper can be found in the slides).
:::

The log-odds of being in a higher wait time category are expected to decrease by 3.4 when the patient arrives by ambulance versus when they arrive by walk-in, while adjusting for all other factors in the model.

::: callout-note
## Part c

Consider output from both the wait time and treatment time models. Use the results from both models to describe the effect of triage level (red = urgent, green = non-urgent) on the wait and treatment times in the ED. Note: red is the baseline level.
:::

The log-odds of being in a higher wait time category are expected to increase by 0.016 when the patient has triage level green versus when they have triage level red, while adjusting for all other factors in the model.

The log-odds of being in a higher treatment time category are expected to decrease by 0.95 when the patient has triage level green versus when they have triage level red, while adjusting for all other factors in the model.

# Exercise 3

Ibanez and Roussel (2022) conducted an experiment to understand the impact of watching a nature documentary on pro-environmental behavior. The researchers randomly assigned the 113 participants to watch an video about architecture in NYC (control) or a video about Yellowstone National Park (treatment). As part of the experiment, participants played a game in which they had an opportunity to donate to an environmental organization. The data set is available in nature.csv in the data folder. We will use the following variables:

-   donation_binary: 1 - participant donated to environmental organization versus 0 - participant did not donate.
-   Age: age in years
-   Gender: Participant’s reported gender
-   Treatment: Urban (T1)” - the control group versus “Nature (T2)” - the treatment group.
-   NEP_high: 1 - score of 4 or higher on the New Ecological Paradigm (NEP) versus 0 - score of less than 4.

See the Introduction and Methods sections of Ibanez and Roussel (2022) for more details about the variables and see the class slides regarding the url for the paper.

```{r}
nature <- read_csv("nature.csv", show_col_types = FALSE)
# https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0275806
nature = nature %>% select(c("donation_binary", "Age", "Gender", "Treatment", "nep_high", "Donation (level)"))
# summary(nature)
```

::: callout-note
## Part a

Figure 2 on pg. 9 of the article visualizes the relationship between donation amount and treatment. Recreate this visualization using your own code. Use the visualization to describe the relationship between donating and the treatment.
:::

```{r}

nature |>
  ggplot(aes(x = `Donation (level)`, fill = Treatment)) +
  geom_bar(position = "dodge", aes(y = after_stat(prop))) +
  scale_x_continuous(breaks = seq(0, 11, by = 1)) +
  theme_bw() +
  labs(
    y = "percentage",
    subtitle = "Range of donation to ENGO, by treatment (Urban (T1) and Nature (T2))"
  ) +
  scale_fill_manual(
    values = c("Urban" = "gray70", "Nature" = "black"),
    name = "Treatment"
  )
```

From this visualization, it seems like participants in the Urban treatment were more likely to donate on average.

::: callout-note
## Part b

Fit a probit regression model using age, gender, treatment, nep_high and the interaction between nep_high and treatment predict the likelihood of donating. (Note: Your model will be similar (but not exactly the same) as the “Likelihood” model in Table 5 on pg. 11.) Display the model.
:::

```{r}

probit_model <- glm(donation_binary ~ Age + Gender + Treatment + nep_high + nep_high * Treatment,
data = nature, family = binomial(link = "probit"))

probit_model |>
tidy() |>
kable(digits = 3)
```

::: callout-note
## Part c

Describe the effect of watching the documentary on the likelihood of donating.
:::

Participants who watched the nature documentary (Nature (T2), 0.191) were more likely to donate compared to those who viewed the urban documentary (T1). However, the high p-value (0.637) suggests that this difference is not statistically significant while accounting for the other variables in the model.

::: callout-note
## Part d

Based on the model, what is the predicted probability of donating for a 20-year old female in the treatment group with a NEP score of 3?
:::

```{r}
data <- data.frame(
Age = 20,
Gender = 0,
Treatment = "NATURE (T2)",
nep_high = 3

)
predict(probit_model, newdata = data, type = "response")
```

According to the model, a 20-year-old female in the treatment group with a NEP score of 3 has an estimated donation probability of about 0.33.
