---
title: "STA 310: Homework 1"
author: "Luke Flyer"
output: 
     pdf_document
font-size: 8px
---

::: callout-important
:::

# Instructions

-   Write all narrative using full sentences. Write all interpretations and conclusions in the context of the data.
-   Be sure all analysis code is displayed in the rendered pdf.
-   If you are fitting a model, display the model output in a neatly formatted table. (The `tidy` and `kable` functions can help!)
-   If you are creating a plot, use clear and informative labels and titles.
-   Render and back up your work reguarly, such as using Github.
-   When you're done, we should be able to render the final version of the Rmd document to fully reproduce your pdf.
-   Upload your pdf to Gradescope. Upload your Rmd, pdf (and any data) to Canvas.

# Exercises

Exercises 1 - 4 are adapted from exercises in Section 1.8 of @roback2021beyond.

## Exercise 1

<!--Adapted from BMLR Ch 1 Ex 1-->

Consider the following scenario:

> Researchers record the number of cricket chirps per minute and temperature during that time. They use linear regression to investigate whether the number of chirps varies with temperature.

a.  Identify the response and predictor variable.

The response variable is the number of chirps per minute, and they predictor variable is temperature.

a.  Write the complete specification of the statistical model.

$Y_i = \beta_0 + \beta_1(Temperature_i) + \epsilon_i$

In this case, $\beta_0$ represents the true intercept—the expected number of cricket chirps per minute at temperature = 0. $\beta_1$ represents the true slope—the expected increase in cricket chirps per minute for every unit increase in temperature, assuming the rate of increase is linear. $\epsilon_i$ represent the deviations of the actual chirps per minute from the expected number predicted by the model.

a.  Write the assumptions for linear regression in the context of the problem.

Linearity: The mean number of cricket chirps per minute is linearly related to temperature.

Independence: Each observation of cricket chirps per minute and temperature is independent and not related to each other.

Normality: The cricket chirps per minute are normally distributed at any given temperature.

Equal variance: Variability of cricket chirps per minute is the same at all temperatures.

## Exercise 2

<!--Adapted from BMLR Ch 1 Ex 2-->

Consider the following scenario:

> A randomized clinical trial investigated postnatal depression and the use of an estrogen patch. Patients were randomly assigned to either use the patch or not. Depression scores were recorded on 6 different visits.

a.  Identify the response and predictor variables.

The response variable is depression score, and the predictor variable is estrogen patch vs. control group.

a.  Identify which model assumption(s) are violated. Briefly explain your choice.

Independence seems to be most violated since depression scores were recorded on 6 different visits, meaning that there are 6 observations for each patient. This would mean that knowing about one of those observations could provide information about another.

## Exercise 3

<!--Adapted from Ch 1 Ex 3-->

Use the [Kentucky Derby case study](https://bookdown.org/roback/bookdown-BeyondMLR/ch-MLRreview.html#review-of-multiple-linear-regression) in Chapter 1 of *Beyond Multiple Linear Regression.*

a.  Consider [Equation (1.3)](https://bookdown.org/roback/bookdown-BeyondMLR/ch-MLRreview.html#multiple-linear-regression-with-two-predictors) in Section 1.6.3. Show why we have to be sure to say "holding year constant", "after adjusting for year", or an equivalent statement, when interpreting $\beta_2$.

We have to say "holding year constant" because the year is another predictor in this linear model. Therefore, $\beta_2$ is the expected difference in winning speeds between slow and fast conditions after accounting for year.

a.  Briefly explain why there is no error (random variation) term $\epsilon_i$ in [Equation (1.4)](https://bookdown.org/roback/bookdown-BeyondMLR/ch-MLRreview.html#multiple-linear-regression-with-an-interaction-term) in Section 1.6.6?

This is because this equation gives us "estimated winning speeds" which does not have to include this error term. The actual observations from the model will have error, but the estimation equation does not.

## Exercise 4

<!--Adapted from Ch 1 Guided Ex 3-->

The data set `kingCountyHouses.csv` in the `data` folder contains data on over 20,000 houses sold in King County, Washington (@kingcounty).

We will use the following variables:

-   `price` = selling price of the house
-   `sqft` = interior square footage

*See Section 1.8 of Beyond Multiple Linear Regression for the full list of variables.*

a.  Fit a linear regression model with `price` as the response variable and `sqft` as the predictor variable (Model 1). Interpret the slope coefficient in terms of the expected change in price when `sqft` increases by 100.

```{r, warning = F, message = F}
library(tidyverse)
library(tidymodels)
library(knitr)
houses <- read_csv("/home/guest/310/homeworks/data/kingCountyHouses.csv")

linear_reg() |>
  set_engine("lm") |>
  fit(price ~ sqft, data = houses) |>
  tidy() |>
  kable()
```

For every increase of 100 square feet, the price of a house is expected to increase by \$28,062.

a.  Fit Model 2, where `logprice` (the natural log of price) is now the response variable and `sqft` is still the predictor variable. How is the `logprice` expected to change when `sqft` increases by 100?

```{r, warning = F, message = F}

linear_reg() |>
  set_engine("lm") |>
  fit(log(price) ~ sqft, data = houses) |>
  tidy() |>
  kable()
```

For every increase of 100 square feet, the log(price) of a house is expected to increase by .0399 log(dollars).

a.  Recall that $log(a) - log(b) = log(\frac{a}{b})$. Use this to derive how the `price` is expected to change when `sqft` increases by 100 based on Model 2.

For every increase of 100 square feet, the price of a house is expected to be multiplied by a factor of $e^{0.0399}$. This is because 0.0399 = $log(\frac{a}{b})$, and we then need to exponentiate both sides to solve for the price ratio of a house with 100 more square feet to the price of the house with 100 fewer.

a.  Fit Model 3, where `price` and `logsqft` (the natural log of sqft) are the response and predictor variables, respectively. How does the price expected to change when sqft increases by 10%? *As a hint, this is the same as multiplying sqft by 1.10.*

```{r, warning = F, message = F}

linear_reg() |>
  set_engine("lm") |>
  fit(price ~ log(sqft), data = houses) |>
  tidy() |>
  kable()
```

```{r}

log(1.1) * 528647.5
```

For a 10% increase in square feet (multiplicative increase of 1.1), we expect the price of a house to increase by \$50385.49.

::: callout-tip
[Click here](https://github.com/STA210-Sp19/supplemental-notes/blob/master/log-transformations.pdf) for notes on interpreting model effects for log-transformed response and/or predictor variables.
:::

## Exercise 5

The goal of this analysis is to use characteristics of 593 colleges and universities in the United States to understand variability in the early career pay, defined as the median salary for alumni with 0 - 5 years of experience. The data was obtained from [TidyTuesday College tuition, diversity, and pay](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-10/readme.md), and was originaly collected from the [PayScale College Salary Report](https://www.payscale.com/college-salary-report/bachelors).

The data set is located in `college-data.csv` in the `data` folder. We will focus on the following variables:

| variable           | class     | description                                                                           |
|:----------------|:----------------|:--------------------------------------|
| name               | character | Name of school                                                                        |
| state_name         | character | state name                                                                            |
| type               | character | Public or private                                                                     |
| early_career_pay   | double    | Median salary for alumni with 0 - 5 years experience (in US dollars)                  |
| stem_percent       | double    | Percent of degrees awarded in science, technology, engineering, or math subjects      |
| out_of_state_total | double    | Total cost for in-state residents in USD (sum of room & board + out of state tuition) |

a.  Visualize the distribution of the response variable `early_career_pay`. Write 1 - 2 observations from the plot.
b.  Visualize the relationship between (i) `early_career_pay` and `type` and (ii) `early_career_pay` and `stem_percent`. Write an observation from each plot.
c.  Below is the specification of the statistical model for this analysis. Fit the model and neatly display the results using 3 digits. Display the 95% confidence interval for the coefficients.

```{=tex}
\begin{align}
early\_career\_pay_{i} = \beta_0 &+ \beta_1~out\_of\_state\_total_{i} + \beta_2 ~ type \\
&+ \beta_3 ~ stem\_percent_{i} + \beta_4 ~ type * stem\_percent_{i} \\ &+ \epsilon_{i}, \hspace{5mm} \text{where } \epsilon_i \sim N(0, \sigma^2)
\end{align}
```
d.  How many degrees of freedom are there in the estimate of the regression standard error $\sigma$?
e.  What is the 95% confidence interval for the amount in which the intercept for public institutions differs from private institutions?

## Exercise 6

Use the analysis from the previous exercise to write a paragraph (\~ 4 - 5 sentences) describing the differences in early career pay based on the institution characteristics. *The summary should be consistent with the results from the previous exercise, comprehensive, answers the primary analysis question, and tells a cohesive story (e.g., a list of interpretations will not receive full credit).*

# Grading

| **Total**             | **50** |
|-----------------------|:------:|
| Ex 1                  |   8    |
| Ex 2                  |   4    |
| Ex 3                  |   7    |
| Ex 4                  |   12   |
| Ex 5                  |   12   |
| Ex 6                  |   4    |
| Workflow & formatting |   3    |

The "Workflow & formatting" grade is to based on the organization of the assignment write up along with the reproducible workflow. This includes having an organized write up with neat and readable headers, code, and narrative, including properly rendered mathematical notation. It also includes having a reproducible Rmd/Quarto document that can be rendered to reproduce the submitted PDF.
